---
title: "Convergence Diagnostic"
author: "Ziang Zhang"
date: "2025-04-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In this tutorial, we examine several diagnostic tools that can be used to assess the convergence of BOSS.

For now, let's assume the following true (log) posterior:

```{r}
library(npreg)
library(ggplot2)
library(aghq)

set.seed(123)
noise_var = 1e-6
function_path <- "./code"
output_path <- "./output/simA1"
data_path <- "./data/simA1"
source(paste0(function_path, "/00_BOSS.R"))

lower = 0; upper = 10

log_prior <- function(x){
  1
}
log_likelihood <- function(x){
  log(x + 1) * (sin(x * 4) + cos(x * 2))
}
eval_once <- function(x){
  log_prior(x) + log_likelihood(x)
}
eval_once_mapped <- function(y){
  eval_once(pnorm(y) * (upper - lower) + lower) + dnorm(y, log = T) + log(upper - lower)
}
x <- seq(0.01,9.99, by = 0.01)
y <- qnorm((x - lower)/(upper - lower))
true_log_norm_constant <- log(integrate(f = function(y) exp(eval_once_mapped(y)), lower = -Inf, upper = Inf)$value)
true_log_post_mapped <- function(y) {eval_once_mapped(y) - true_log_norm_constant}
```


```{r}
plot((true_log_post_mapped(y)) ~ y, type = "l", cex.lab = 1.5, cex.axis = 1.5, 
     xlab = "y", ylab = "Log density", lwd = 2, col = "blue")
true_log_post <- function(x) {true_log_post_mapped(qnorm((x - lower)/(upper - lower))) - dnorm(qnorm((x - lower)/(upper - lower)), log = T) - log(upper - lower)}
integrate(function(x) exp(true_log_post(x)), lower = 0, upper = 10)
plot(exp(true_log_post(x)) ~ x, type = "l", cex.lab = 1.5, cex.axis = 1.5, 
     xlab = "x", ylab = "Posterior density", lwd = 2, col = "blue")
```


## KL Divergence

Let $f_t$ and $f_{t-j}$ be the corresponding surrogate density at time $t$ and $t-j$, respectively.
We can compute the KL divergence between $f_t$ and $f_{t-j}$ as follows:
$$Q_t = KL(f_t,f_{t-j}) = \int \log \frac{f_t(x)}{f_{t-j}(x)}f_{t}(x)dx.$$

For one-dimensional problems, this can be done efficiently through numerical integration.
For higher-dimensional problems, sampling-based methods can be used to approximate the KL divergence.


```{r}
result_ad <- BOSS(
  func = eval_once, initial_design = 5,
  update_step = 5, max_iter = 30,
  opt.lengthscale.grid = 100, opt.grid = 1000,
  delta = 0.01, noise_var = noise_var,
  lower = lower, upper = upper,
  verbose = 0,
  KL_iter_check = 1, KL_check_warmup = 5, KL_eps = 0, criterion = "KL"
)

plot(result_ad$KL_result$KL ~ result_ad$KL_result$i,
     xlab = "Iteration", ylab = "KL statistic",
     main = "KL statistic over iterations", 
     log = "y",
     pch = 19, col = "blue")
```


Based on the KL divergence, it seems like the algorithm has converged around 30 iterations.


## KS Statistics


The Kolmogorov-Smirnov (KS) statistic measures the maximum difference between the cumulative distribution functions (CDFs) $F_t$ and $F_{t-j}$ of the surrogate densities $f_t$ and $f_{t-j}$, respectively.
Specifically, for one dimensional problems, the KS statistic is defined as:
$$Q_t = \max_x |F_t(x) - F_{t-j}(x)|.$$


```{r}
result_ad <- BOSS(
  func = eval_once, initial_design = 5,
  update_step = 5, max_iter = 30,
  opt.lengthscale.grid = 100, opt.grid = 1000,
  delta = 0.01, noise_var = noise_var,
  lower = lower, upper = upper,
  verbose = 0,
  KS_iter_check = 1, KS_check_warmup = 5, KS_eps = 0, criterion = "KS"
)

plot(result_ad$KS_result$KS ~ result_ad$KS_result$i,
     xlab = "Iteration", ylab = "KS statistic",
     main = "KS statistic over iterations", 
     log = "y",
     pch = 19, col = "blue")
```


Based on the KS statistic, the conclusion is similar to that of the KL divergence. The KS statistics is very close to 0 after 30 iterations, indicating that the algorithm has likely converged.



## Modal Convergence


For higher-dimensional problems, computing KL divergence is computationally more intensive due to the need for numerical integration or sampling-based methods. 

Due to Bernstein-Von-Mises theorem, when the sample size is large, the majority of the posterior mass is concentrated around the mode.
Thus, as an empirical heuristic, we can check the convergence of the modal behavior of the surrogate density. Although this method is less rigorous than KL divergence or KS statistic, it is computationally efficient and can be used as a quick diagnostic check for convergence.

For the modal behavior, we compute the average $k$-nearest neighbor distance between the current optimizer and its neighboring design points, as well as the relative change of the hessian (trace) at the current optimizer. 

Specifically, define the current optimizer as $\boldsymbol{\theta}_t^*$ and the average $k$-nearest neighbor distance of design points to $\boldsymbol{\theta}_t^*$ as $d_{t,k}$. 
Additionally, let $H(\boldsymbol{\theta}_t^*)$ be the negative Hessian at $\boldsymbol{\theta}_t^*$ based on the GP surrogate and write $S_{t} = \sqrt{\mathrm{Tr}(H^{-1}(\boldsymbol{\theta}_t^*))}$. We can assess modal convergence by considering

1. $D_{t,k} = d_{t,k}/S_{t,k}$;
2. $H_{t} = \frac{S_{t} - S_{t-1}}{S_{t-1}}$.

We deem the algorithm to have reached convergence when $Q_t =\max\{D_{t,k}, H_{t}\} < \varepsilon$ for some user-defined $\varepsilon > 0$. 
Heuristically, $S_{t}$ can be regarded as the size of the posterior standard deviation for sufficiently large $t$. 
$D_{t,k}$ thus gives a measure of how close the current optimizer is to the existing design points relative to the size of the posterior standard deviation. 
$H_{t}$, on the other hand, gives a rough measure on the relative change of the size of the posterior standard deviation.


```{r}
result_ad <- BOSS(
  func = eval_once, initial_design = 5,
  update_step = 5, max_iter = 30,
  opt.lengthscale.grid = 100, opt.grid = 1000,
  delta = 0.01, noise_var = noise_var,
  lower = lower, upper = upper,
  verbose = 0,
  modal_iter_check = 1, modal_check_warmup = 10, modal_k.nn = 5, modal_eps = 0, criterion = "modal"
)

plot(result_ad$modal_result$modal ~ result_ad$modal_result$i,
     xlab = "Iteration", ylab = "Modal statistic",
     main = "Modal statistic over iterations", 
     log = "y",
     pch = 19, col = "blue")
```


Again, the modal statistic converges to 0 after 30 iterations, indicating that the algorithm has likely converged.






