---
title: "problem_with_opt"
author: "Ziang Zhang"
date: "2025-04-27"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Here I am showing an example where `opt` is not working as expected for AF. 

Let's load the example in Simulation 4.1, and take a look at the "oracle" posterior density. 

```{r eval=FALSE}
library(BayesGP)
library(tidyverse)
library(npreg)
function_path <- "./code"
output_path <- "./output/sim1"
data_path <- "./data/sim1"
source(paste0(function_path, "/00_BOSS.R"))

lower = 0.5; upper = 4.5; a = 1.5; noise_var = 1e-6
### Simulate data:
set.seed(123)
n <- 100
x <- runif(n = n, min = 0, max = 10)
true_func <- function(x, true_alpha = 1.5){1 + 
    0.5 * cos((2*pi*x)/true_alpha) - 1.3 * sin((2*pi*x)/true_alpha) +
    1.1 * cos((4*pi*x)/true_alpha) + 0.3 * sin((4*pi*x)/true_alpha)}
log_mu <- true_func(x) + rnorm(n, sd = 2)
y <- rpois(n = n, lambda = exp(log_mu))
data <- data.frame(y = y, x = x, indx = 1:n, log_mu = log_mu)
log_prior <- function(alpha){
  dnorm(x = alpha, mean = 3, log = T, sd = 0.5)
}
eval_once <- function(alpha){
  a_fit <- (2*pi)/alpha
  x <- data$x
  data$cosx <- cos(a_fit * x)
  data$sinx <- sin(a_fit * x)
  data$cos2x <- cos(2*a_fit * x)
  data$sin2x <- sin(2*a_fit * x)
  mod <- model_fit(formula = y ~ cosx + sinx + cos2x + sin2x + f(x = indx, model = "IID", 
                                                                 sd.prior = list(param = 1)),
                   data = data, method = "aghq", family = "Poisson", aghq_k = 4
  )
  (mod$mod$normalized_posterior$lognormconst) + log_prior(alpha)
}
surrogate <- function(xvalue, data_to_smooth, choice_cov) {
  predict_gp(
    data = data_to_smooth,
    x_pred = matrix(xvalue, ncol = 1),
    choice_cov = choice_cov,
    noise_var = noise_var
  )$mean
}
x_vals <- seq(lower, upper, by = 0.005)
load(paste0(output_path, "/exact_grid_result.rda"))
plot(x = exact_grid_result$x, y = exact_grid_result$pos, col = "black", cex = 0.5, type = "l",
     xlab = "x", ylab = "density", main = "Posterior Density (Exact Grid)", lwd = 2)
abline(v = exact_grid_result$x[which.max(exact_grid_result$exact_vals)], col = "green", lty = "dashed")
abline(v = exact_grid_result$x[which.max(exact_grid_result$exact_vals)], col = "blue", lty = "dashed")
abline(v = a, col = "purple", lty = "dashed")
grid()
```


```{r}
library(BayesGP)
library(tidyverse)
library(npreg)
function_path <- "./code"
output_path <- "./output/sim2"
data_path <- "./data/sim2"
source(paste0(function_path, "/00_BOSS.R"))

set.seed(123)
lower = 0; upper = 10
x_grid <- seq(lower, upper, length.out = 1000)
noise_var <- 1e-6

func_generator <- function(f1, f2) {
  func <- function(x, a) {
    sapply(x, function(xi) {
      if (xi <= a) {
        f1(xi)
      } else {
        f2(xi) - f2(a) + f1(a)
      }
    })
  }
  func
}
my_func <- func_generator(f1 = function(x) x*log(x^2 + 1), f2 = function(x) 3.3*x)

# Simulate observations from a regression model with a piecewise linear function
simulate_observation <- function(a, func, x_grid, measurement = 3) {
  # Generate n random x values between 0 and 1
  x <- rep(x_grid, each = measurement)
  n <- length(x)
  
  # Initialize y
  y <- numeric(n)
  
  # Loop through each x to compute the corresponding y value based on the piecewise function
  fx <- func(x = x, a = a)
  
  # Add random noise e from a standard normal distribution
  e <- rnorm(n, mean = 0, sd = 0.3)
  y <- fx + e
  
  return(data.frame(x, y))
}

# Knot value
a <- 6.5
data <- simulate_observation(a = a, func = my_func, x_grid = x_grid, measurement = 1)

lower = 0.1; upper = 9.9
eval_once <- function(alpha){
  a_fit <- alpha
  data$x1 <- ifelse(data$x <= a_fit, data$x, a_fit)
  data$x2 <- ifelse(data$x > a_fit, (data$x - a_fit), 0)
  mod <- model_fit(formula = y ~ f(x1, model = "IWP", order = 2, sd.prior = list(param = 1, h = 1), initial_location = 0) + f(x2, model = "IWP", order = 2, sd.prior = list(param = 1, h = 1), initial_location = 0), 
                   data = data, method = "aghq", family = "Gaussian", aghq_k = 3
  )
  (mod$mod$normalized_posterior$lognormconst)
}
# Apply smoothing to the result
surrogate <- function(xvalue, data_to_smooth){
  data_to_smooth$y <- data_to_smooth$y - mean(data_to_smooth$y)
  predict(ss(x = as.numeric(data_to_smooth$x), y = data_to_smooth$y, df = length(unique(as.numeric(data_to_smooth$x))), m = 2, all.knots = TRUE), x = xvalue)$y
}
load(paste0(output_path, "/exact_grid_result.rda"))
load(paste0(output_path, "/exact_grid_result_smooth.rda"))
plot(exact_grid_result$x, exact_grid_result$pos, type = "l", col = "red", xlab = "x (0-10)", ylab = "density", main = "Posterior")
abline(v = a, col = "purple")
grid()
```



## Using grid-based optimization of AF

First, let's see what we get when we use the grid-based optimization of AF. 


```{r}
objective_func <- eval_once
max_iter <- 20
result_grid <- BOSS(func = objective_func, update_step = 5, max_iter = (max_iter-5),
                  delta = 0.01, noise_var = noise_var,
                  opt.lengthscale.grid = 100, opt.grid = 2000,
                  lower = lower, upper = upper,
                  criterion = "modal", verbose = 0,
                  modal_iter_check = 1, modal_check_warmup = 5, modal_k.nn = 5, modal_eps = 0,
                  initial_design = 5)

data_to_smooth <- result_grid$result
data_to_smooth$y <- data_to_smooth$y - mean(data_to_smooth$y)

ff <- list()
ff$fn <- function(x) as.numeric(surrogate(x, data_to_smooth = data_to_smooth))
x_vals <- (seq(from = lower, to = upper, length.out = 1000) - lower)/(upper - lower)
fn_vals <- sapply(x_vals, ff$fn)
dx <- diff(x_vals)
# Compute the trapezoidal areas and sum them up
integral_approx <- sum(0.5 * (exp(fn_vals[-1]) + exp(fn_vals[-length(x_vals)])) * dx)
post_x <- data.frame(y = x_vals, pos = exp(fn_vals)/integral_approx)
post_x_final <- data.frame(x = (lower + x_vals*(upper - lower)), pos = post_x$pos /(upper - lower))

ggplot() +
    geom_line(data = post_x_final, aes(x = x, y = pos), color = "red", size = 1) +
    geom_line(data = exact_grid_result_smooth, aes(x = x, y = pos), color = "black", size = 0.5, linetype = "dashed") +
    ggtitle(paste0("Comparison Posterior Density: B = ", max_iter)) +
    xlab("Value") +
    ylab("Density") +
    coord_cartesian(ylim = c(0,10), xlim = c(5,8)) + 
    theme(text=element_text(size=10)) +
    theme_minimal() 
```

Let's also take a look at the convergence of modal statistics:

```{r}
plot(result_grid$modal_result$modal_max_diff ~ result_grid$modal_result$i, xlab = "Iteration", ylab = "Modal Statistics", main = "Convergence of Modal Statistics")
```

As well as the placement of design points:

```{r}
plot(result_grid$result$x_original, xlab = "iteration", ylab = "Design Points", ylim = c(lower,upper))
```



## Using opt-based optimization of AF

Then, let's see what we get when we use the opt-based optimization of AF. 

```{r}
objective_func <- eval_once
result_opt <- BOSS(func = objective_func, update_step = 5, max_iter = (max_iter-5),
                  delta = 0.01, noise_var = noise_var,
                  lower = lower, upper = upper,
                  opt.lengthscale.grid = 100, 
                  criterion = "modal", verbose = 3,
                  modal_iter_check = 1, modal_check_warmup = 5, modal_k.nn = 5, modal_eps = 0,
                  initial_design = 5)

data_to_smooth <- result_opt$result
data_to_smooth$y <- data_to_smooth$y - mean(data_to_smooth$y)

ff <- list()
ff$fn <- function(x) as.numeric(surrogate(x, data_to_smooth = data_to_smooth))
x_vals <- (seq(from = lower, to = upper, length.out = 1000) - lower)/(upper - lower)
fn_vals <- sapply(x_vals, ff$fn)
dx <- diff(x_vals)
# Compute the trapezoidal areas and sum them up
integral_approx <- sum(0.5 * (exp(fn_vals[-1]) + exp(fn_vals[-length(x_vals)])) * dx)
post_x <- data.frame(y = x_vals, pos = exp(fn_vals)/integral_approx)
post_x_final <- data.frame(x = (lower + x_vals*(upper - lower)), pos = post_x$pos /(upper - lower))

ggplot() +
    geom_line(data = post_x_final, aes(x = x, y = pos), color = "red", size = 1) +
    geom_line(data = exact_grid_result_smooth, aes(x = x, y = pos), color = "black", size = 0.5, linetype = "dashed") +
    ggtitle(paste0("Comparison Posterior Density: B = ", max_iter)) +
    xlab("Value") +
    ylab("Density") +
    coord_cartesian(ylim = c(0,10), xlim = c(5,8)) + 
    theme(text=element_text(size=10)) +
    theme_minimal() 
```


Let's also take a look at the convergence of modal statistics:

```{r}
plot(result_opt$modal_result$modal_max_diff ~ result_opt$modal_result$i, xlab = "Iteration", ylab = "Modal Statistics", main = "Convergence of Modal Statistics")
```

As well as the placement of design points:

```{r}
plot(result_opt$result$x_original, xlab = "iteration", ylab = "Design Points", ylim = c(lower,upper))
```
