---
title: "Simulation 3: Two-dimensional Non-linear Regression"
author: "Dayi Li"
date: "2025-04-19"
output: 
  workflowr::wflow_html:
    code_folding: show
editor_options:
  chunk_output_type: console
---

## Data

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(tikzDevice)
library(rstan)
library(INLA)
library(inlabru)
library(modeest)

function_path <- "./code"
output_path <- "./output/sim3"
data_path <- "./data/sim3"
source(paste0(function_path, "/00_BOSS.R"))
```

Consider the following non-linear regression model:

\begin{align*}
y_i \mid \log(\rho_i) &\overset{ind}{\sim}\mathcal{N}(\log\rho(r_i), \sigma^2), \\
\log\rho(r_i) &= \log\rho_0 - \gamma\log\left\{1 + (r_i/R)^\beta\right\}.
\end{align*}

We simulate $n = 200$ data points based on the above model with $\rho_0 = 10$, $R = 2$, $\beta = 2$, $\gamma = -2.5$, and $\sigma = 0.5$. The inferential goal is the nuisance parameters $R$ and $\beta$.

```{r}
r <- seq(0, 20, length.out = 50)
beta <- 10
a <- 2
b <- 2
c <- -2.5

set.seed(1234)
Ir <- beta*(1 + (r/a)^b)^c
lr <- log(Ir) + rnorm(length(r), 0, 0.5)

data <- data.frame(r, lr)

ggplot(data, aes(r, lr)) + geom_point() + ylab('y')
```

## inlabru

We first run `inlabru` to to fit the model. We set the following priors for the parameters:

\begin{align*}
\rho_0 \sim \mathcal{N}(0, 1000), \ & R \sim \mathrm{Unif}(0.1, 5), \\
\beta \sim \mathrm{Unif}(0.1, 4),  \ \gamma \sim \mathcal{N}(0, & 1000), \ \sigma^2 \sim \mathrm{Inv-Gamma}(1, 10^{-5}).
\end{align*}

```{r}
a_fun <- function(u){
  qunif(pnorm(u), 0.1, 5)
}

b_fun <- function(u){
  qunif(pnorm(u), 0.1, 4)
}

cmp <- ~ a(1, model="linear", mean.linear=0, prec.linear=1) +
  b(1, model="linear", mean.linear=0, prec.linear=1) + 
  c(1) + Intercept(1)

form <- lr ~ Intercept + c*log(1 + (r/a_fun(a))^b_fun(b))

fit <- bru(cmp, formula = form, data = data, family = 'gaussian')
```

## BOSS

Now let's run BOSS. 

### BOSS-modal

We first specify the (unnormalized) log-posterior for $(R,\beta)$. Note that for this specific problem, the unnormalized log-posterior has a closed-form expression:

```{r}
# specify the objective function for BOSS: unnormalized log posterior of (R, beta)
eval_func <- function(par, x = r, y = lr){
  a <- par[1]
  b <- par[2]
  n <- length(r)
  
  X <- matrix(cbind(rep(1, n), log(1 + (r/a)^b)), ncol = 2)
  Vb <- solve(t(X) %*% X + diag(1/1000, 2))
  P <- diag(n) - X %*% Vb %*% t(X)
  
  mlik <- log(det(Vb))/2 - log(1000) + lgamma((n+1)/2) - (n+1)/2*log(1e-5 + t(y) %*% P %*% y/2) - 
    n/2*log(pi) -5*log(10)
  
  return(mlik)
}
```

### BOSS-modal

Next, we run the BOSS algorithm where the stopping criteria is based on the convergence of the posterior mode. Specifically, we check the modal convergence every BO iteration, and consider the convergence statistics of the average $5$ nearest neighbor distance around the current mode.

```{r, eval = F}
set.seed(1234)
res_opt_modal <- BOSS(eval_func, criterion = 'modal', update_step = 5, max_iter = 100, D = 2,
                      lower = rep(0.1, 2), upper = c(5, 4),
                      noise_var = 1e-6,
                      modal_iter_check = 1,  modal_check_warmup = 20, modal_k.nn = 5,
                      modal_eps = 0.01,
                      initial_design = 5, delta = 0.01^2,
                      optim.n = 5, optim.max.iter = 100)

save(res_opt_modal, file = paste0(output_path, "/BOSS_modal_sim3.rda"))
```

The above BOSS algorithm with `modal` convergence criterion converged in $91$ iterations.

Let's check the convergence diagnostic plot:

```{r}
load(paste0(output_path, "/BOSS_modal_sim3.rda"))

plot(res_opt_modal$modal_result$modal_max_diff ~ res_opt_modal$modal_result$i, type = 'o', xlab = 'Iteration', ylab = 'Modal Stats')
```


## MCMC

Lastly, we implement the MCMC-based method using `stan` to obtain the oracle. 

```{r, eval = F}
set.seed(1234)
MCMC_fit <- stan(
  file = "code/nlreg.stan",  # Stan program
  data = list(x = r, y = lr, N = length(r)),    # named list of data
  chains = 4,             # number of Markov chains
  warmup = 3000,          # number of warmup iterations per chain
  iter = 30000,            # total number of iterations per chain
  cores = 4,              # number of cores (could use one per chain)
  algorithm = 'NUTS')

# thin the samples fo plotting
MCMC_samp <- as.data.frame(MCMC_fit)
#MCMC_samp_thin <- MCMC_samp[seq(1, 76000, by = 8),]
save(MCMC_samp, file = paste0(output_path, "/MCMC_sim3.rda"))
```

## Results Comparison

### Marginal Posterior Distributions

We first compare the marginal posterior distributions of $R$ and $\beta$ from `inlabru`, BOSS, and MCMC.

```{r}
# inlabru marginal samples
set.seed(1234)
inla.samples.a <- a_fun(inla.rmarginal(108000, fit$marginals.fixed$a))
inla.samples.b <- b_fun(inla.rmarginal(108000, fit$marginals.fixed$b))
```

```{r eval = F}
# BOSS-modal marginal samples
data_to_smooth <- list()
unique_data <- unique(data.frame(x = res_opt_modal$result$x, y = res_opt_modal$result$y))
data_to_smooth$x <- as.matrix(dplyr::select(unique_data, -y))
data_to_smooth$y <- (unique_data$y - mean(unique_data$y))

square_exp_cov <- square_exp_cov_generator_nd(length_scale = res_opt_modal$length_scale, signal_var = res_opt_modal$signal_var)

surrogate <- function(xvalue, data_to_smooth, cov){
  predict_gp(data_to_smooth, x_pred = xvalue, choice_cov = cov, noise_var = 1e-6)$mean
}

ff <- list()
ff$fn <- function(x) as.numeric(surrogate(x, data_to_smooth = data_to_smooth, cov = square_exp_cov))

x.1 <- (seq(from = 0.1, to = 5, length.out = 500) - 0.1)/4.9
x.2 <- (seq(from = 0.1, to = 4, length.out = 500) - 0.1)/3.9
x_vals <- expand.grid(x.1, x.2)
names(x_vals) <- c('x.1','x.2')
x_original <- t(t(x_vals)*(c(5, 4) - c(0.1, 0.1)) + c(0.1, 0.1)) 

fn_vals <- apply(x_vals, 1, function(x) ff$fn(x = matrix(x, ncol = 2))) + mean(unique_data$y)
fn_true <- apply(x_original, 1, function(x) eval_func(x))
# normalize
lognormal_const <- log(sum(exp(fn_vals))*0.0098*0.0078)
lognormal_const_true <- log(sum(exp(fn_true))*0.0098*0.0078)

post_x_modal <- data.frame(x_original, pos = exp(fn_vals - lognormal_const))
post_x_true <- data.frame(x_original, pos = exp(fn_true - lognormal_const_true))

save(post_x_modal, file = paste0(output_path, "/BOSS_post.rda"))
save(post_x_true, file = paste0(output_path, "/true_post.rda"))

```


```{r}
load(paste0(output_path, "/BOSS_post.rda"))
load(paste0(output_path, "/true_post.rda"))

dx <- unique(post_x_modal$x.1)[2] - unique(post_x_modal$x.1)[1]
dy <- unique(post_x_modal$x.2)[2] - unique(post_x_modal$x.2)[1]
set.seed(123456)
sample_idx <- rmultinom(1:250000, size = 108000, prob = post_x_modal$pos)
sample_x_modal <- data.frame(post_x_modal, n = sample_idx)
sample_idx_true <- rmultinom(1:250000, size = 108000, prob = post_x_true$pos)
sample_x_true <- data.frame(post_x_true, n = sample_idx_true)

samples_BOSS_modal <- data.frame(do.call(rbind, apply(sample_x_modal, 1, function(x) cbind(runif(x[4], x[1], x[1]+dx), runif(x[4], x[2], x[2] + dy)))))
samples_true <- data.frame(do.call(rbind, apply(sample_x_true, 1, function(x) cbind(runif(x[4], x[1], x[1]+dx), runif(x[4], x[2], x[2] + dy)))))

# MCMC marginal samples
load(paste0(output_path, "/MCMC_sim3.rda"))

# Combine all samples together

R_marginal <- data.frame(R = c(inla.samples.a, samples_BOSS_modal[,1], MCMC_samp$a,
                               samples_true[,1]), 
                         method = rep(c('inlabru', 'BOSS',  'MCMC', 'Grid (Oracle)'),  
                                      c(length(inla.samples.a), 
                                        length(samples_BOSS_modal[,1]), 
                                        length(MCMC_samp$a), length(samples_true[,1]))))

beta_marginal <- data.frame(beta = c(inla.samples.b, samples_BOSS_modal[,2], MCMC_samp$b,
                                     samples_true[,2]), 
                            method = rep(c('inlabru', 'BOSS', 'MCMC', 'Grid (Oracle)'),  
                                         c(length(inla.samples.b),
                                           length(samples_BOSS_modal[,2]),  
                                           length(MCMC_samp$b), length(samples_true[,2]))))
```

Plot the marginal posterior densities

```{r}
#tikz(file = "R_marginal.tex", standAlone=T, width = 4, height = 3)
ggplot(R_marginal, aes(R)) + geom_density(aes(color = method), show_guide = F) +
  stat_density(aes(x = R, colour = method),
                  geom="line",position="identity") + theme_minimal() + xlab('$R$')
#dev.off()
#system('pdflatex R_marginal.tex')
```

```{r}
#tikz(file = "beta_marginal.tex", standAlone=T, width = 4, height = 3)
ggplot(beta_marginal, aes(beta)) + geom_density(aes(color = method), show_guide = F) +
  stat_density(aes(x = beta, colour = method),
                  geom="line",position="identity") + theme_minimal() + xlab('$\\beta$')
#dev.off()
#system('pdflatex beta_marginal.tex')
```

### Joint Posterior Distribution

We now compare the results of the posterior distributions from `inlabru`, modal-based BOSS, and AGHQ-based BOSS, and MCMC.

#### `inlabru` joint posterior distribution:

```{r, warning = F}
# get joint posterior of (R, beta) from inlabru
joint_samp <- inla.posterior.sample(10000, fit, selection = list(a = 1, b = 1), seed = 12345)
joint_samp <- do.call('rbind', lapply(joint_samp, function(x) matrix(x$latent, ncol = 2)))

inla.joint.samps <- data.frame(a = a_fun(joint_samp[,1]), b = b_fun(joint_samp[,2]))

# plot joint posterior of (R, beta) from inlabru
#tikz(file = "joint_post_R_beta_inlabru.tex", standAlone=T, width = 4, height = 3)
ggplot(inla.joint.samps, aes(a, b)) + stat_density_2d(
  geom = "raster",
  aes(fill = after_stat(density)), n = 300,
  contour = FALSE) +
  geom_point(data = data.frame(a = a_fun(fit$summary.fixed$mode[1]), b = b_fun(fit$summary.fixed$mode[2])), color = 'red', shape = 1, size =0.5) + 
  geom_point(data = data.frame(a = 2, b = 2), color = 'orange', size =0.5) +
  coord_fixed() + scale_fill_viridis_c(name = 'Density') + theme_minimal() + xlab('$R$') + ylab('$\\beta$') + xlim(c(0.1, 5)) + ylim(c(0.1, 4))
#dev.off()
#system('pdflatex joint_post_R_beta_inlabru.tex')
```

#### BOSS joint posterior distribution:

```{r}
# plot joint posterior of (R, beta) from BOSS
#tikz(file = "joint_post_R_beta.tex", standAlone=T, width = 4, height = 3)
ggplot(post_x_modal, aes(x.1,x.2)) + geom_raster(aes(fill = (pos))) + 
  geom_point(data = data.frame(x.1 = post_x_modal$x.1[which.max(post_x_modal$pos)], x.2 = post_x_modal$x.2[which.max(post_x_modal$pos)]), color = 'red', shape = 1, size =0.5) +
  geom_point(data = data.frame(x.1 = 2, x.2 = 2), color = 'orange', size =0.5) + coord_fixed() + scale_fill_viridis_c(name = 'Density') + theme_minimal() + xlab('$R$') + ylab('$\\beta$')
#dev.off()
#system('pdflatex joint_post_R_beta.tex')
```

#### MCMC joint posterior distribution:

```{r, warning=F}
#tikz(file = "joint_post_R_beta_MCMC.tex", standAlone=T, width = 4, height = 3)
ggplot(MCMC_samp, aes(a, b)) + stat_density_2d(
  geom = "raster",
  aes(fill = after_stat(density)), n = 300,
  contour = FALSE) + 
  geom_point(data = data.frame(a = post_x_modal$x.1[which.max(post_x_modal$pos)], b = post_x_modal$x.2[which.max(post_x_modal$pos)]), color = 'red', shape = 1, size =0.5) +
  geom_point(data = data.frame(a = 2, b = 2), color = 'orange', size =0.5) + coord_fixed() + scale_fill_viridis_c(name = 'Density') + theme_minimal() + xlab('$R$') + ylab('$\\beta$') + xlim(c(0.1, 5)) + ylim(c(0.1, 4))
#dev.off()
#system('pdflatex joint_post_R_beta_MCMC.tex')
```

#### True joint posterior distribution (grid):

```{r}
# plot joint posterior of (R, beta) from BOSS
#tikz(file = "joint_post_R_beta_oracle.tex", standAlone=T, width = 4, height = 3)
ggplot(post_x_true, aes(x.1,x.2)) + geom_raster(aes(fill = (pos))) + 
  geom_point(data = data.frame(x.1 = post_x_true$x.1[which.max(post_x_true$pos)], x.2 = post_x_true$x.2[which.max(post_x_true$pos)]), color = 'red', shape = 1, size =0.5) +
  geom_point(data = data.frame(x.1 = 2, x.2 = 2), color = 'orange', size =0.5) + coord_fixed() + scale_fill_viridis_c(name = 'Density') + theme_minimal() + xlab('$R$') + ylab('$\\beta$')
#dev.off()
#system('pdflatex joint_post_R_beta_oracle.tex')
```


From the above results, it is clear that BOSS is much better at depicting the joint posterior distribution than `inlabru`. The joint distribution from `inlabru` is simply the product of the marginal distribution, which completely ignores the more complex structures in the joint posterior. Interestingly though, BOSS is also much better than HMC. The posterior samples from HMC is significantly oversampling from the tail regions. This is likely due to highly warped geometry of the posterior and the un-optimized step number and step size used in the NUTS sampler.


## Does Starting Design Points Matter?

We here check if BOSS is robust towards the starting design points. We run BOSS for $20$ times with different initial design points and run until the modal statistics have reached below $\epsilon =0.01$. 

```{r, eval = F}
res_list <- vector('list', 20)

for (i in 1:20) {
  set.seed(i)
  res_list[[i]] <- BOSS(eval_func, criterion = 'modal', update_step = 5, max_iter = 100, D = 2,
                        lower = rep(0.1, 2), upper = c(5, 4),
                        noise_var = 1e-6,
                        modal_iter_check = 1,  modal_check_warmup = 20, modal_k.nn = 5,
                        modal_eps = 0.01,
                        initial_design = 5, delta = 0.01^2,
                        optim.n = 5, optim.max.iter = 100)
}

save(res_list, file = paste0(output_path, "/BOSS_robust_sim3.rda"))
```


```{r, eval = F}
load(paste0(output_path, "/BOSS_robust_sim3.rda"))

sample_list <- vector('list', 20)

for(i in 1:20){
  # BOSS-modal marginal samples
  data_to_smooth <- list()
  unique_data <- unique(data.frame(x = res_list[[i]]$result$x, y = res_list[[i]]$result$y))
  data_to_smooth$x <- as.matrix(dplyr::select(unique_data, -y))
  data_to_smooth$y <- (unique_data$y - mean(unique_data$y))
  
  square_exp_cov <- square_exp_cov_generator_nd(length_scale = res_list[[i]]$length_scale, signal_var = res_list[[i]]$signal_var)
  
  surrogate <- function(xvalue, data_to_smooth, cov){
    predict_gp(data_to_smooth, x_pred = xvalue, choice_cov = cov, noise_var = 1e-6)$mean
  }
  
  ff <- list()
  ff$fn <- function(x) as.numeric(surrogate(x, data_to_smooth = data_to_smooth, cov = square_exp_cov))
  
  x.1 <- (seq(from = 0.1, to = 5, length.out = 300) - 0.1)/4.9
  x.2 <- (seq(from = 0.1, to = 4, length.out = 300) - 0.1)/3.9
  x_vals <- expand.grid(x.1, x.2)
  names(x_vals) <- c('x.1','x.2')
  x_original <- t(t(x_vals)*(c(5, 4) - c(0.1, 0.1)) + c(0.1, 0.1)) 
  
  fn_vals <- apply(x_vals, 1, function(x) ff$fn(x = matrix(x, ncol = 2))) + mean(unique_data$y)
  # normalize
  lognormal_const <- log(sum(exp(fn_vals))*0.0098*0.0078*25/9)
  post_x_modal <- data.frame(x_original, pos = exp(fn_vals - lognormal_const))
  
  dx <- unique(post_x_modal$x.1)[2] - unique(post_x_modal$x.1)[1]
  dy <- unique(post_x_modal$x.2)[2] - unique(post_x_modal$x.2)[1]
  set.seed(123456)
  sample_idx <- rmultinom(1:90000, size = 49500, prob = post_x_modal$pos)
  sample_x_modal <- data.frame(post_x_modal, n = sample_idx)
  
  sample_list[[i]] <- data.frame(do.call(rbind, apply(sample_x_modal, 1, function(x) cbind(runif(x[4], x[1], x[1]+dx), runif(x[4], x[2], x[2] + dy)))))
  
}

save(sample_list, file = paste0(output_path, "/BOSS_sample_small_n_sim3.rda"))
```

```{r}
load(paste0(output_path, "/BOSS_sample_small_n_sim3.rda"))

p_R <- ggplot(samples_true, aes(x = X1)) + theme_minimal() + xlab('$R$')

for (i in 1:20) {
  p_R <- p_R + geom_density(data = sample_list[[i]], aes(x = X1), color = 'red', alpha = 0.1)
}

p_R <- p_R + geom_density(color = 'black', size = 1.5) 

#tikz(file = "marginal_R_robust.tex", standAlone=T, width = 4, height = 3)
p_R
#dev.off()
#system('pdflatex marginal_R_robust.tex')

p_b <- ggplot(samples_true, aes(x = X2)) + theme_minimal() + xlab('$\\beta$')

for (i in 1:20) {
  p_b <- p_b + geom_density(data = sample_list[[i]], aes(x = X2), color = 'red', alpha = 0.1)
}

p_b <- p_b + geom_density(color = 'black', size = 1.5) 

#tikz(file = "marginal_beta_robust.tex", standAlone=T, width = 4, height = 3)
p_b
#dev.off()
#system('pdflatex marginal_beta_robust.tex')
```

In the above, black density comes from the grid-based oracle posterior while red ones come from BOSS. We can see that BOSS is highly robust to the initial design points selection.




