---
title: "Explore Complex Posterior Functions"
author: "Ziang Zhang"
date: "2025-04-21"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

## Introduction

In this example, we assess the accuracy of the proposed BOSS algorithm for approximating (un-normalized) posterior functions with different complexities. 

We consider a conditioning parameter $\alpha$ with support $\Omega = [0,10]$, and assume its unnormalized log posteriors are defined respectively as $f(\alpha) = \alpha \sin(\alpha)$, $\log(\alpha+1)\sin(2\alpha) - \alpha \cos(2\alpha)$,
and $\log(\alpha + 1) (\sin(4\alpha) + \cos(2\alpha))$ for simple, medium, and hard settings.

In the simple setting, the log posterior has two local modes and its corresponding posterior is close to uni-modal. 
In the medium scenario, the log posterior has three local modes, with the corresponding posterior that is close to bi-modal. 
In the hard scenario, the log posterior has seven local modes and the posterior is close to tri-modal.
The proposed BOSS algorithm is then applied with different numbers of BO iterations $B$ with five initial values that are equally placed from 0 to 10.

```{r}
library(npreg)
library(ggplot2)
library(aghq)

set.seed(123)
noise_var = 1e-6
function_path <- "./code"
output_path <- "./output/simA1"
data_path <- "./data/simA1"
source(paste0(function_path, "/00_BOSS.R"))

surrogate <- function(xvalue, data_to_smooth){
  predict(ss(x = as.numeric(data_to_smooth$x), y = data_to_smooth$y, df = length(unique(as.numeric(data_to_smooth$x))), m = 2, all.knots = TRUE), x = xvalue)$y
}
lower = 0
upper = 10

integrate_aghq <- function(f, k = 100, startingvalue = 0){
  ff <- list(fn = f, gr = function(x) numDeriv::grad(f, x), he = function(x) numDeriv::hessian(f, x))
  aghq(ff = ff, k = k, startingvalue = startingvalue)$normalized_posterior$lognormconst
}

#### Compute the KL distance:
Compute_KL <- function(x, logpx, logqx){
  dx <- diff(x)
  left <- c(0,dx)
  right <- c(dx,0)
  0.5 * sum(left * (logpx - logqx) * exp(logpx)) + 0.5 * sum(right * (logpx - logqx) * exp(logpx))
}

#### Compute the KS distance:
Compute_KS <- function(x, qx, px){
  dx <- c(diff(x),0)
  max(abs(cumsum(qx * dx) - cumsum(px * dx)))
}
```


## Easy Example:

Define the log posterior function:

```{r}
log_prior <- function(x){
  1
}
log_likelihood <- function(x){
  x*sin(x)
}
eval_once <- function(x){
  log_prior(x) + log_likelihood(x)
}
eval_once_mapped <- function(y){
  eval_once(pnorm(y) * (upper - lower) + lower) + dnorm(y, log = T) + log(upper - lower)
}
x <- seq(0.01,9.99, by = 0.01)
y <- qnorm((x - lower)/(upper - lower))
true_log_norm_constant <- integrate_aghq(f = function(y) eval_once_mapped(y))
true_log_post_mapped <- function(y) {eval_once_mapped(y) - true_log_norm_constant}
plot((true_log_post_mapped(y)) ~ y, type = "l", cex.lab = 1.5, cex.axis = 1.5, 
     xlab = "y", ylab = "log density", lwd = 2, col = "blue")
true_log_post <- function(x) {true_log_post_mapped(qnorm((x - lower)/(upper - lower))) - dnorm(qnorm((x - lower)/(upper - lower)), log = T) - log(upper - lower)}
integrate(function(x) exp(true_log_post(x)), lower = 0, upper = 10)
```


Let's run BOSS with the above log posterior function:

```{r}
objective_func <- eval_once
eval_num <- seq(5, 100, by = 5)
```

```{r eval=FALSE}
result_ad <- BOSS(
  func = eval_once, initial_design = 5,
  update_step = 5, max_iter = (max(eval_num) - 5),
  opt.lengthscale.grid = 100, opt.grid = 1000,
  delta = 0.01, noise_var = noise_var,
  lower = lower, upper = upper,
  verbose = 0,
  modal_iter_check = 5, modal_check_warmup = 10, modal_k.nn = 5, modal_eps = 0, criterion = "modal"
)
saveRDS(result_ad, file = paste0(data_path, "/result_ad_easy.rds"))
```


```{r eval=FALSE}
BO_result_list <- list()
BO_result_original_list <- list()
result_ad <- readRDS(file = paste0(data_path, "/result_ad_easy.rds"))
for (i in 1:length(eval_num)) {
  eval_number <- eval_num[i]

  result_ad_selected <- list(x = result_ad$result$x[1:eval_number, ],
                             x_original = result_ad$result$x_original[1:eval_number, ],
                             y = result_ad$result$y[1:eval_number])
  
  data_to_smooth <- result_ad_selected
  BO_result_original_list[[i]] <- data_to_smooth

  ff <- list()
  ff$fn <- function(y) as.numeric(surrogate(pnorm(y), data_to_smooth = data_to_smooth) + dnorm(y, log = TRUE))
  fn_vals <- sapply(y, ff$fn)

  lognormal_const <- integrate_aghq(f = ff$fn)
  post_y <- data.frame(y = y, pos = exp(fn_vals - lognormal_const))
  post_x <- data.frame(x = pnorm(post_y$y) * (upper - lower) + lower, post = (post_y$pos / dnorm(post_y$y))/(upper - lower) )
  
  BO_result_list[[i]] <- post_x
}
saveRDS(BO_result_list, file = paste0(data_path, "/BO_result_list_easy.rds"))
saveRDS(BO_result_original_list, file = paste0(data_path, "/BO_result_original_list_easy.rds"))
```

### Some illustrations{.tabset .tabset-pills}

Here are some illustrations of the approximated posterior from BOSS with different numbers of BO iterations $B$ (red: approximated posterior, black: true posterior). 


```{r}
BO_result_list <- readRDS(file = paste0(data_path, "/BO_result_list_easy.rds"))
BO_result_original_list <- readRDS(file = paste0(data_path, "/BO_result_original_list_easy.rds"))
```

#### B = `r eval_num[1]`

```{r fig.width=6, fig.height=8}
to_plot_data <- BO_result_list[[1]]
to_plot_data$logpos <- log(to_plot_data$pos)
to_plot_data_obs <- BO_result_original_list[[1]]
y_min <- -15; y_max <- 0
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0), mfrow = c(2, 1))
plot(to_plot_data$logpos ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Log Post", xlab = expression(alpha),
       ylim = c(y_min, y_max), cex.lab = 1.5, cex.axis = 1.5)
lines((true_log_post(x)) ~ x, lwd = 1)
y_offset <- 0.03 * (y_max - y_min) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}

plot(exp(to_plot_data$logpos) ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Post", xlab = expression(alpha), ylim = c(0,2), cex.lab = 1.5, cex.axis = 1.5)
lines(exp(true_log_post(x)) ~ x, lwd = 1)
y_min <- -0.03
y_offset <- 0.01 * (2) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}
```

#### B = `r eval_num[2]`

```{r fig.width=6, fig.height=8}
to_plot_data <- BO_result_list[[2]]
to_plot_data$logpos <- log(to_plot_data$pos)
to_plot_data_obs <- BO_result_original_list[[2]]
y_min <- -15; y_max <- 0

mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0), mfrow = c(2, 1))
plot(to_plot_data$logpos ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Log Post", xlab = expression(alpha),
       ylim = c(y_min, y_max), cex.lab = 1.5, cex.axis = 1.5)
lines((true_log_post(x)) ~ x, lwd = 1)
y_offset <- 0.03 * (y_max - y_min) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}

plot(exp(to_plot_data$logpos) ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Post", xlab = expression(alpha), ylim = c(0,2), cex.lab = 1.5, cex.axis = 1.5)
lines(exp(true_log_post(x)) ~ x, lwd = 1)
y_min <- -0.03
y_offset <- 0.01 * (2) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}
```

#### B = `r eval_num[6]`

```{r fig.width=6, fig.height=8}
to_plot_data <- BO_result_list[[6]]
to_plot_data$logpos <- log(to_plot_data$pos)
to_plot_data_obs <- BO_result_original_list[[6]]
y_min <- -15; y_max <- 0

mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0), mfrow = c(2, 1))
plot(to_plot_data$logpos ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Log Post", xlab = expression(alpha),
       ylim = c(y_min, y_max), cex.lab = 1.5, cex.axis = 1.5)
lines((true_log_post(x)) ~ x, lwd = 1)
y_offset <- 0.03 * (y_max - y_min) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}

plot(exp(to_plot_data$logpos) ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Post", xlab = expression(alpha), ylim = c(0,2), cex.lab = 1.5, cex.axis = 1.5)
lines(exp(true_log_post(x)) ~ x, lwd = 1)
y_min <- -0.03
y_offset <- 0.01 * (2) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}
```

#### B = `r eval_num[20]`

```{r fig.width=6, fig.height=8}
to_plot_data <- BO_result_list[[20]]
to_plot_data$logpos <- log(to_plot_data$pos)
to_plot_data_obs <- BO_result_original_list[[20]]
y_min <- -15; y_max <- 0

mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0), mfrow = c(2, 1))
plot(to_plot_data$logpos ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Log Post", xlab = expression(alpha),
       ylim = c(y_min, y_max), cex.lab = 1.5, cex.axis = 1.5)
lines((true_log_post(x)) ~ x, lwd = 1)
y_offset <- 0.03 * (y_max - y_min) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}

plot(exp(to_plot_data$logpos) ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Post", xlab = expression(alpha), ylim = c(0,2), cex.lab = 1.5, cex.axis = 1.5)
lines(exp(true_log_post(x)) ~ x, lwd = 1)
y_min <- -0.03
y_offset <- 0.01 * (2) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}
```


```{r eval=FALSE, include=FALSE}
for (i in 1:length(eval_num)) {
  eval_number <- eval_num[i]

  to_plot_data <- BO_result_list[[i]]
  to_plot_data$logpos <- log(to_plot_data$pos)
  to_plot_data_obs <- BO_result_original_list[[i]]
  y_min <- -15
  y_max <- 0
  png(filename = paste0(output_path, "/easy/", "BOSS_approxi_B_", eval_number, "_easy.png"), width = 800, height = 800)
  mar.default <- c(5,4,4,2)
  par(mar = mar.default + c(0, 1, 0, 0))
  plot(to_plot_data$logpos ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 3,
       ylab = "Log Post", xlab = expression(alpha),
       ylim = c(y_min, y_max), cex.lab = 2.5, cex.axis = 2.5)
  lines((true_log_post(x)) ~ x, lwd = 1)
  y_offset <- 0.03 * (y_max - y_min) # adjust offset as needed
  for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
  }
  dev.off()
  
  png(filename = paste0(output_path, "/easy/", "pos_BOSS_approxi_B_", eval_number, "_easy.png"), width = 800, height = 800)
  mar.default <- c(5,4,4,2)
  par(mar = mar.default + c(0, 1, 0, 0))
  plot(exp(to_plot_data$logpos) ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Post", xlab = expression(alpha), ylim = c(0,2), cex.lab = 2.5, cex.axis = 2.5)
  lines(exp(true_log_post(x)) ~ x, lwd = 1)
  y_min <- -0.03
  y_offset <- 0.01 * (2) # adjust offset as needed
  for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
  }
  dev.off()
  
}
```



### KL and KS:

Let's visualize the KL and KS divergence between the true posterior and the approximated posterior from BOSS. 

```{r}
KL_vec <- c()
for (i in 1:length(eval_num)) {
  KL_vec[i] <- Compute_KL(x = x, logpx = true_log_post(x), logqx = log(BO_result_list[[i]]$pos))
}
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0))
plot((KL_vec) ~ eval_num, type = "o", ylab = "KL", xlab = "eval number: B", cex.lab = 1.5, cex.axis = 1.5)

KS_vec <- c()
for (i in 1:length(eval_num)) {
  KS_vec[i] <- Compute_KS(x = x, px = exp(true_log_post(x)), qx = BO_result_list[[i]]$pos)
}
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0))
plot((KS_vec) ~ eval_num, type = "o", ylab = "KS", xlab = "eval number: B", ylim = c(0,1), cex.lab = 1.5, cex.axis = 1.5)
```

```{r eval=FALSE, include=FALSE}
KL_vec <- c()
for (i in 1:length(eval_num)) {
  KL_vec[i] <- Compute_KL(x = x, logpx = true_log_post(x), logqx = log(BO_result_list[[i]]$pos))
}
png(filename = paste0(output_path, "/easy/kl_compare_easy.png"), height = 800, width = 800)
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0))
plot((KL_vec) ~ eval_num, type = "o", ylab = "KL", xlab = "eval number: B", cex.lab = 2.5, cex.axis = 2.5)
dev.off()

KS_vec <- c()
for (i in 1:length(eval_num)) {
  KS_vec[i] <- Compute_KS(x = x, px = exp(true_log_post(x)), qx = BO_result_list[[i]]$pos)
}
png(filename = paste0(output_path, "/easy/ks_compare_easy.png"), height = 800, width = 800)
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0))
plot((KS_vec) ~ eval_num, type = "o", ylab = "KS", xlab = "eval number: B", ylim = c(0,1), cex.lab = 2.5, cex.axis = 2.5)
dev.off()

png(filename = paste0(output_path, "/easy/log_ks_compare_easy.png"), height = 800, width = 800)
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0))
plot(log(KS_vec) ~ eval_num, type = "o", ylab = "Log KS", xlab = "eval number: B", ylim = c(-10,0), cex.lab = 2.5, cex.axis = 2.5)
dev.off()
```

Based on the KL and KS divergence, we can the approximated posterior from BOSS is indistinguishable from the true posterior after 10 iterations in the simple example.



## Medium Example:

Define the log posterior function:

```{r}
log_prior <- function(x){
  1
}
log_likelihood <- function(x){
  log(x+1)*sin(x*2) - x*cos(x*2)
}
eval_once <- function(x){
  log_prior(x) + log_likelihood(x)
}
eval_once_mapped <- function(y){
  eval_once(pnorm(y) * (upper - lower) + lower) + dnorm(y, log = T) + log(upper - lower)
}
x <- seq(0.01,9.99, by = 0.01)
y <- qnorm((x - lower)/(upper - lower))
true_log_norm_constant <- integrate_aghq(f = function(y) eval_once_mapped(y), k = 100)
true_log_post_mapped <- function(y) {eval_once_mapped(y) - true_log_norm_constant}
plot((true_log_post_mapped(y)) ~ y, type = "l", cex.lab = 1.5, cex.axis = 1.5, 
     xlab = "y", ylab = "log density", lwd = 2, col = "blue")
true_log_post <- function(x) {true_log_post_mapped(qnorm((x - lower)/(upper - lower))) - dnorm(qnorm((x - lower)/(upper - lower)), log = T) - log(upper - lower)}
integrate(function(x) exp(true_log_post(x)), lower = 0, upper = 10)
```

Let's run BOSS with the above log posterior function:

```{r}
objective_func <- eval_once
```

```{r eval=FALSE}
result_ad <- BOSS(
  func = objective_func, initial_design = 5,
  update_step = 5, max_iter = (max(eval_num) - 5),
  opt.lengthscale.grid = 100, opt.grid = 1000,
  delta = 0.01, noise_var = noise_var,
  lower = lower, upper = upper,
  modal_iter_check = 5, modal_check_warmup = 10, modal_k.nn = 5, modal_eps = 0, criterion = "modal"
)
saveRDS(result_ad, file = paste0(data_path, "/result_ad_med.rds"))
```

```{r eval=FALSE}
BO_result_list <- list()
BO_result_original_list <- list()
result_ad <- readRDS(file = paste0(data_path, "/result_ad_med.rds"))
for (i in 1:length(eval_num)) {
  eval_number <- eval_num[i]

  result_ad_selected <- list(x = result_ad$result$x[1:eval_number, ],
                             x_original = result_ad$result$x_original[1:eval_number, ],
                             y = result_ad$result$y[1:eval_number])
  
  data_to_smooth <- result_ad_selected
  BO_result_original_list[[i]] <- data_to_smooth

  ff <- list()
  ff$fn <- function(y) as.numeric(surrogate(pnorm(y), data_to_smooth = data_to_smooth) + dnorm(y, log = TRUE))
  fn_vals <- sapply(y, ff$fn)

  lognormal_const <- integrate_aghq(f = ff$fn)
  post_y <- data.frame(y = y, pos = exp(fn_vals - lognormal_const))
  post_x <- data.frame(x = pnorm(post_y$y) * (upper - lower) + lower, post = (post_y$pos / dnorm(post_y$y))/(upper - lower) )
  
  BO_result_list[[i]] <- post_x
}
saveRDS(BO_result_list, file = paste0(data_path, "/BO_result_list_med.rds"))
saveRDS(BO_result_original_list, file = paste0(data_path, "/BO_result_original_list_med.rds"))
```

### Some illustrations{.tabset .tabset-pills}

Here are some illustrations of the approximated posterior from BOSS with different numbers of BO iterations $B$ (red: approximated posterior, black: true posterior). 

```{r}
BO_result_list <- readRDS(file = paste0(data_path, "/BO_result_list_med.rds"))
BO_result_original_list <- readRDS(file = paste0(data_path, "/BO_result_original_list_med.rds"))
```


#### B = `r eval_num[1]`

```{r fig.width=6, fig.height=8}
to_plot_data <- BO_result_list[[1]]
to_plot_data$logpos <- log(to_plot_data$pos)
to_plot_data_obs <- BO_result_original_list[[1]]
y_min <- -20; y_max <- 5
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0), mfrow = c(2, 1))
plot(to_plot_data$logpos ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Log Post", xlab = expression(alpha),
       ylim = c(y_min, y_max), cex.lab = 1.5, cex.axis = 1.5)
lines((true_log_post(x)) ~ x, lwd = 1)
y_offset <- 0.03 * (y_max - y_min) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}

plot(exp(to_plot_data$logpos) ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Post", xlab = expression(alpha), ylim = c(0, 2.5), cex.lab = 1.5, cex.axis = 1.5)
lines(exp(true_log_post(x)) ~ x, lwd = 1)
y_min <- -0.03
y_offset <- 0.01 * (2) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}
```

#### B = `r eval_num[2]`

```{r fig.width=6, fig.height=8}
to_plot_data <- BO_result_list[[2]]
to_plot_data$logpos <- log(to_plot_data$pos)
to_plot_data_obs <- BO_result_original_list[[2]]
y_min <- -20; y_max <- 5

mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0), mfrow = c(2, 1))
plot(to_plot_data$logpos ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Log Post", xlab = expression(alpha),
       ylim = c(y_min, y_max), cex.lab = 1.5, cex.axis = 1.5)
lines((true_log_post(x)) ~ x, lwd = 1)
y_offset <- 0.03 * (y_max - y_min) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}

plot(exp(to_plot_data$logpos) ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Post", xlab = expression(alpha), ylim = c(0, 2.5), cex.lab = 1.5, cex.axis = 1.5)
lines(exp(true_log_post(x)) ~ x, lwd = 1)
y_min <- -0.03
y_offset <- 0.01 * (2) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}
```

#### B = `r eval_num[6]`

```{r fig.width=6, fig.height=8}
to_plot_data <- BO_result_list[[6]]
to_plot_data$logpos <- log(to_plot_data$pos)
to_plot_data_obs <- BO_result_original_list[[6]]
y_min <- -20; y_max <- 5

mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0), mfrow = c(2, 1))
plot(to_plot_data$logpos ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Log Post", xlab = expression(alpha),
       ylim = c(y_min, y_max), cex.lab = 1.5, cex.axis = 1.5)
lines((true_log_post(x)) ~ x, lwd = 1)
y_offset <- 0.03 * (y_max - y_min) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}

plot(exp(to_plot_data$logpos) ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Post", xlab = expression(alpha), ylim = c(0, 2.5), cex.lab = 1.5, cex.axis = 1.5)
lines(exp(true_log_post(x)) ~ x, lwd = 1)
y_min <- -0.03
y_offset <- 0.01 * (2) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}
```

#### B = `r eval_num[20]`

```{r fig.width=6, fig.height=8}
to_plot_data <- BO_result_list[[20]]
to_plot_data$logpos <- log(to_plot_data$pos)
to_plot_data_obs <- BO_result_original_list[[20]]
y_min <- -20; y_max <- 5

mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0), mfrow = c(2, 1))
plot(to_plot_data$logpos ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Log Post", xlab = expression(alpha),
       ylim = c(y_min, y_max), cex.lab = 1.5, cex.axis = 1.5)
lines((true_log_post(x)) ~ x, lwd = 1)
y_offset <- 0.03 * (y_max - y_min) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}

plot(exp(to_plot_data$logpos) ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Post", xlab = expression(alpha), ylim = c(0, 2.5), cex.lab = 1.5, cex.axis = 1.5)
lines(exp(true_log_post(x)) ~ x, lwd = 1)
y_min <- -0.03
y_offset <- 0.01 * (2) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}
```


```{r eval=FALSE, include=FALSE}
for (i in 1:length(eval_num)) {
  eval_number <- eval_num[i]

  to_plot_data <- BO_result_list[[i]]
  to_plot_data$logpos <- log(to_plot_data$pos)
  to_plot_data_obs <- BO_result_original_list[[i]]
  y_min <- -20
  y_max <- 5
  png(filename = paste0(output_path, "/medium/", "BOSS_approxi_B_", eval_number, "_med.png"), width = 800, height = 800)
  mar.default <- c(5,4,4,2)
  par(mar = mar.default + c(0, 1, 0, 0))
  plot(to_plot_data$logpos ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 3,
       ylab = "Log Post", xlab = expression(alpha),
       ylim = c(y_min, y_max), cex.lab = 2.5, cex.axis = 2.5)
  lines((true_log_post(x)) ~ x, lwd = 1)
  y_offset <- 0.03 * (y_max - y_min) # adjust offset as needed
  for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
  }
  dev.off()
  
  png(filename = paste0(output_path, "/medium/", "pos_BOSS_approxi_B_", eval_number, "_med.png"), width = 800, height = 800)
  mar.default <- c(5,4,4,2)
  par(mar = mar.default + c(0, 1, 0, 0))
  plot(exp(to_plot_data$logpos) ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Post", xlab = expression(alpha), ylim = c(0, 2.5), cex.lab = 2.5, cex.axis = 2.5)
  lines(exp(true_log_post(x)) ~ x, lwd = 1)
  y_min <- -0.03
  y_offset <- 0.01 * (2) # adjust offset as needed
  for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
  }
  dev.off()
  
}
```


### KL and KS:

Let's visualize the KL and KS divergence between the true posterior and the approximated posterior from BOSS. 


```{r}
KL_vec <- c()
for (i in 1:length(eval_num)) {
  KL_vec[i] <- Compute_KL(x = x, logpx = true_log_post(x), logqx = log(BO_result_list[[i]]$pos))
}
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0))
plot((KL_vec) ~ eval_num, type = "o", ylab = "KL", xlab = "eval number: B", cex.lab = 1.5, cex.axis = 1.5)

KS_vec <- c()
for (i in 1:length(eval_num)) {
  KS_vec[i] <- Compute_KS(x = x, px = exp(true_log_post(x)), qx = BO_result_list[[i]]$pos)
}
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0))
plot((KS_vec) ~ eval_num, type = "o", ylab = "KS", xlab = "eval number: B", ylim = c(0,1), cex.lab = 1.5, cex.axis = 1.5)
```

```{r eval=FALSE, include=FALSE}
KL_vec <- c()
for (i in 1:length(eval_num)) {
  KL_vec[i] <- Compute_KL(x = x, logpx = true_log_post(x), logqx = log(BO_result_list[[i]]$pos))
}
png(filename = paste0(output_path, "/medium/kl_compare_med.png"), height = 800, width = 800)
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0))
plot((KL_vec) ~ eval_num, type = "o", ylab = "KL", xlab = "eval number: B", cex.lab = 2.5, cex.axis = 2.5)
dev.off()

KS_vec <- c()
for (i in 1:length(eval_num)) {
  KS_vec[i] <- Compute_KS(x = x, px = exp(true_log_post(x)), qx = BO_result_list[[i]]$pos)
}
png(filename = paste0(output_path, "/medium/ks_compare_med.png"), height = 800, width = 800)
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0))
plot((KS_vec) ~ eval_num, type = "o", ylab = "KS", xlab = "eval number: B", ylim = c(0,1), cex.lab = 2.5, cex.axis = 2.5)
dev.off()

png(filename = paste0(output_path, "/medium/log_ks_compare_med.png"), height = 800, width = 800)
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0))
plot(log(KS_vec) ~ eval_num, type = "o", ylab = "Log KS", xlab = "eval number: B", ylim = c(-10,0), cex.lab = 2.5, cex.axis = 2.5)
dev.off()
```

Based on the KL and KS divergence, we can the approximated posterior from BOSS is indistinguishable from the true posterior after 20 iterations of BO in the medium example. 

## Hard Example:

Define the log posterior function:

```{r}
log_prior <- function(x){
  1
}
log_likelihood <- function(x){
  log(x + 1) * (sin(x * 4) + cos(x * 2))
}
eval_once <- function(x){
  log_prior(x) + log_likelihood(x)
}
eval_once_mapped <- function(y){
  eval_once(pnorm(y) * (upper - lower) + lower) + dnorm(y, log = T) + log(upper - lower)
}
x <- seq(0.01,9.99, by = 0.01)
y <- qnorm((x - lower)/(upper - lower))
true_log_norm_constant <- log(integrate(f = function(y) exp(eval_once_mapped(y)), lower = -Inf, upper = Inf)$value)
true_log_post_mapped <- function(y) {eval_once_mapped(y) - true_log_norm_constant}
plot((true_log_post_mapped(y)) ~ y, type = "l", cex.lab = 1.5, cex.axis = 1.5, 
     xlab = "y", ylab = "log density", lwd = 2, col = "blue")
true_log_post <- function(x) {true_log_post_mapped(qnorm((x - lower)/(upper - lower))) - dnorm(qnorm((x - lower)/(upper - lower)), log = T) - log(upper - lower)}
integrate(function(x) exp(true_log_post(x)), lower = 0, upper = 10)
```

Let's run BOSS with the above log posterior function:

```{r}
objective_func <- eval_once
```

```{r eval=FALSE}
result_ad <- BOSS(
  func = objective_func, initial_design = 5,
  update_step = 5, max_iter = (max(eval_num) - 5),
  opt.lengthscale.grid = 100, opt.grid = 1000,
  delta = 0.01, noise_var = noise_var,
  lower = lower, upper = upper,
  AGHQ_iter_check = Inf, AGHQ_eps = 0
)
saveRDS(result_ad, file = paste0(data_path, "/result_ad_hard.rds"))
```

```{r eval=FALSE}
BO_result_list <- list()
BO_result_original_list <- list()
result_ad <- readRDS(file = paste0(data_path, "/result_ad_hard.rds"))
for (i in 1:length(eval_num)) {
  eval_number <- eval_num[i]

  result_ad_selected <- list(x = result_ad$result$x[1:eval_number, ],
                             x_original = result_ad$result$x_original[1:eval_number, ],
                             y = result_ad$result$y[1:eval_number])
  
  data_to_smooth <- result_ad_selected
  BO_result_original_list[[i]] <- data_to_smooth

  ff <- list()
  ff$fn <- function(y) as.numeric(surrogate(pnorm(y), data_to_smooth = data_to_smooth) + dnorm(y, log = TRUE))
  fn_vals <- sapply(y, ff$fn)

  lognormal_const <- log(integrate(f = function(y) exp(ff$fn(y)), lower = -Inf, upper = Inf)$value)
  post_y <- data.frame(y = y, pos = exp(fn_vals - lognormal_const))
  post_x <- data.frame(x = pnorm(post_y$y) * (upper - lower) + lower, post = (post_y$pos / dnorm(post_y$y))/(upper - lower) )
  
  BO_result_list[[i]] <- post_x
}
saveRDS(BO_result_list, file = paste0(data_path, "/BO_result_list_hard.rds"))
saveRDS(BO_result_original_list, file = paste0(data_path, "/BO_result_original_list_hard.rds"))
```

### Some illustrations{.tabset .tabset-pills}

Here are some illustrations of the approximated posterior from BOSS with different numbers of BO iterations $B$ (red: approximated posterior, black: true posterior). 

```{r}
BO_result_list <- readRDS(file = paste0(data_path, "/BO_result_list_hard.rds"))
BO_result_original_list <- readRDS(file = paste0(data_path, "/BO_result_original_list_hard.rds"))
```


#### B = `r eval_num[1]`

```{r fig.width=6, fig.height=8}
to_plot_data <- BO_result_list[[1]]
to_plot_data$logpos <- log(to_plot_data$pos)
to_plot_data_obs <- BO_result_original_list[[1]]
y_min <- -20; y_max <- 5
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0), mfrow = c(2, 1))
plot(to_plot_data$logpos ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Log Post", xlab = expression(alpha),
       ylim = c(y_min, y_max), cex.lab = 1.5, cex.axis = 1.5)
lines((true_log_post(x)) ~ x, lwd = 1)
y_offset <- 0.03 * (y_max - y_min) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}

plot(exp(to_plot_data$logpos) ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Post", xlab = expression(alpha), ylim = c(0,2), cex.lab = 1.5, cex.axis = 1.5)
lines(exp(true_log_post(x)) ~ x, lwd = 1)
y_min <- -0.03
y_offset <- 0.01 * (2) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}
```

#### B = `r eval_num[2]`

```{r fig.width=6, fig.height=8}
to_plot_data <- BO_result_list[[2]]
to_plot_data$logpos <- log(to_plot_data$pos)
to_plot_data_obs <- BO_result_original_list[[2]]
y_min <- -20; y_max <- 5
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0), mfrow = c(2, 1))
plot(to_plot_data$logpos ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Log Post", xlab = expression(alpha),
       ylim = c(y_min, y_max), cex.lab = 1.5, cex.axis = 1.5)
lines((true_log_post(x)) ~ x, lwd = 1)
y_offset <- 0.03 * (y_max - y_min) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}

plot(exp(to_plot_data$logpos) ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Post", xlab = expression(alpha), ylim = c(0,2), cex.lab = 1.5, cex.axis = 1.5)
lines(exp(true_log_post(x)) ~ x, lwd = 1)
y_min <- -0.03
y_offset <- 0.01 * (2) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}
```

#### B = `r eval_num[6]`

```{r fig.width=6, fig.height=8}
to_plot_data <- BO_result_list[[6]]
to_plot_data$logpos <- log(to_plot_data$pos)
to_plot_data_obs <- BO_result_original_list[[6]]
y_min <- -20; y_max <- 5
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0), mfrow = c(2, 1))
plot(to_plot_data$logpos ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Log Post", xlab = expression(alpha),
       ylim = c(y_min, y_max), cex.lab = 1.5, cex.axis = 1.5)
lines((true_log_post(x)) ~ x, lwd = 1)
y_offset <- 0.03 * (y_max - y_min) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}

plot(exp(to_plot_data$logpos) ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Post", xlab = expression(alpha), ylim = c(0,2), cex.lab = 1.5, cex.axis = 1.5)
lines(exp(true_log_post(x)) ~ x, lwd = 1)
y_min <- -0.03
y_offset <- 0.01 * (2) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}
```

#### B = `r eval_num[20]`

```{r fig.width=6, fig.height=8}
to_plot_data <- BO_result_list[[20]]
to_plot_data$logpos <- log(to_plot_data$pos)
to_plot_data_obs <- BO_result_original_list[[20]]
y_min <- -20; y_max <- 5
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0), mfrow = c(2, 1))
plot(to_plot_data$logpos ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Log Post", xlab = expression(alpha),
       ylim = c(y_min, y_max), cex.lab = 1.5, cex.axis = 1.5)
lines((true_log_post(x)) ~ x, lwd = 1)
y_offset <- 0.03 * (y_max - y_min) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}
y_min <- -20; y_max <- 5
plot(exp(to_plot_data$logpos) ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Post", xlab = expression(alpha), ylim = c(0,2), cex.lab = 1.5, cex.axis = 1.5)
lines(exp(true_log_post(x)) ~ x, lwd = 1)
y_min <- -0.03
y_offset <- 0.01 * (2) # adjust offset as needed
for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
}
```


```{r eval=FALSE, include=FALSE}
for (i in 1:length(eval_num)) {
  eval_number <- eval_num[i]
  to_plot_data <- BO_result_list[[i]]
  to_plot_data$logpos <- log(to_plot_data$pos)
  to_plot_data_obs <- BO_result_original_list[[i]]
  y_min <- -20
  y_max <- 5
  
  png(filename = paste0(output_path, "/hard/", "BOSS_approxi_B_", eval_number, "_hard.png"), width = 800, height = 800)
  mar.default <- c(5,4,4,2)
  par(mar = mar.default + c(0, 1, 0, 0))
  plot(to_plot_data$logpos ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 3,
       ylab = "Log Post", xlab = expression(alpha),
       ylim = c(y_min, y_max), cex.lab = 2.5, cex.axis = 2.5)
  lines((true_log_post(x)) ~ x, lwd = 1)
  y_offset <- 0.03 * (y_max - y_min) # adjust offset as needed
  for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
  }
  dev.off()
  
  y_min <- 0
  y_max <- 2.5
  png(filename = paste0(output_path, "/hard/", "pos_BOSS_approxi_B_", eval_number, "_hard.png"), width = 800, height = 800)
  mar.default <- c(5,4,4,2)
  par(mar = mar.default + c(0, 1, 0, 0))
  plot(exp(to_plot_data$logpos) ~ to_plot_data$x, 
       type = "l", lty = "dashed", col = "red", lwd = 2,
       ylab = "Post", xlab = expression(alpha), ylim = c(y_min, y_max), cex.lab = 2.5, cex.axis = 2.5)
  lines(exp(true_log_post(x)) ~ x, lwd = 1)
  y_offset <- 0.01 * (y_max - y_min) # adjust offset as needed
  for(x_val in to_plot_data_obs$x_original) {
    segments(x_val, y_min, x_val, y_min - y_offset, col = "red")
  }
  dev.off()
}
```



### Compute KL and KS:

Let's visualize the KL and KS divergence between the true posterior and the approximated posterior from BOSS. 

```{r}
KL_vec <- c()
for (i in 1:length(eval_num)) {
  KL_vec[i] <- Compute_KL(x = x, logpx = true_log_post(x), logqx = log(BO_result_list[[i]]$pos))
}

mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0))
plot((KL_vec) ~ eval_num, type = "o", ylab = "KL", xlab = "eval number: B", cex.lab = 1.5, cex.axis = 1.5)

KS_vec <- c()
for (i in 1:length(eval_num)) {
  KS_vec[i] <- Compute_KS(x = x, px = exp(true_log_post(x)), qx = BO_result_list[[i]]$pos)
}
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0))
plot((KS_vec) ~ eval_num, type = "o", ylab = "KS", xlab = "eval number: B", ylim = c(0,1), cex.lab = 1.5, cex.axis = 1.5)
```


```{r eval=FALSE, include=FALSE}
KL_vec <- c()
for (i in 1:length(eval_num)) {
  KL_vec[i] <- Compute_KL(x = x, logpx = true_log_post(x), logqx = log(BO_result_list[[i]]$pos))
}

png(filename = paste0(output_path, "/hard/kl_compare_hard.png"), height = 800, width = 800)
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0))
plot((KL_vec) ~ eval_num, type = "o", ylab = "KL", xlab = "eval number: B", cex.lab = 2.5, cex.axis = 2.5)
dev.off()

KS_vec <- c()
for (i in 1:length(eval_num)) {
  KS_vec[i] <- Compute_KS(x = x, px = exp(true_log_post(x)), qx = BO_result_list[[i]]$pos)
}

png(filename = paste0(output_path, "/hard/ks_compare_hard.png"), height = 800, width = 800)
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0))
plot((KS_vec) ~ eval_num, type = "o", ylab = "KS", xlab = "eval number: B", ylim = c(0,1), cex.lab = 2.5, cex.axis = 2.5)
dev.off()

png(filename = paste0(output_path, "/hard/log_ks_compare_hard.png"), height = 800, width = 800)
mar.default <- c(5,4,4,2)
par(mar = mar.default + c(0, 1, 0, 0))
plot(log(KS_vec) ~ eval_num, type = "o", ylab = "Log KS", xlab = "eval number: B", ylim = c(-10,0), cex.lab = 2.5, cex.axis = 2.5)
dev.off()
```

Based on the KL and KS divergence, we can the approximated posterior from BOSS is indistinguishable from the true posterior after around 20 iterations in this hard example.



