---
title: "Detecting Ultra-Diffuse Galaxies"
author: "Dayi Li"
date: "2025-04-21"
output: 
  workflowr::wflow_html:
    code_folding: show
editor_options:
  chunk_output_type: console
---

## Introduction

For this example, we look at the real-world application of using log-Gaussian Cox process (LGCP) to detect a class of extremely faint Ultra-Diffuse Galaxies (UDGs). The main idea is to detect the clustering signals of their constituent star clusters. The data being modeled is thus the point pattern of these star clusters.

The model we consider is as follow:



```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(spatial)
library(raster)
library(zoo)
library(spatstat)
library(excursions)
library(INLA)
library(inlabru)
library(spatialEco)
library(RColorBrewer)
library(R.utils)
library(doParallel)
library(foreach)
library(tikzDevice)
library(aghq)
library(coda)

function_path <- "./code"
output_path <- "./output/UDG"
data_path <- "./data/UDG"
source(paste0(function_path, "/00_BOSS.R"))
```

## Data

We first load the point pattern data we will model, and construct the necessary covariates for modeling.

```{r message=F, warning=F}
# read in data
v_acs <- read_csv(paste0(data_path,'/v10acs_GC.csv'))

# specify study regions
X <- c(1, 4095, 4210, 220)/4300*76
Y <- c(1, 100.5, 4300, 4042)/4300*76
region <- Polygon(cbind(X,Y))
region <- SpatialPolygons(list(Polygons(list(region),'region')))
region <- SpatialPolygonsDataFrame(region, data.frame(id = region@polygons[[1]]@ID, row.names = region@polygons[[1]]@ID))
plot(region)
points(v_acs[,3:4]/4300*76)

# Construct Spatial covariate that captures GC intensity in a bright normal galaxy

grid <- makegrid(region, n = 20000)
grid <- SpatialPoints(grid, proj4string = CRS(proj4string(region)))
grid <- crop(grid, region)

gal1 <- as.data.frame(grid)

names(gal1) <- c('x', 'y')

gal1$z <- (gal1$x - 3278/4300*76)^2/1.4^2 + (gal1$y - 4073/4300*76)^2

coordinates(gal1) <- ~x+y

gridded(gal1) <- T

# construct non-linear functional for covariate
f.gal1 <- function(x, y) {
  spp <- SpatialPoints(data.frame(x = x, y = y), proj4string = fm_sp_get_crs(gal1))
  proj4string(spp) <- fm_sp_get_crs(gal1)
  
  v <- over(spp, gal1)
  if (any(is.na(v$z))) {
    v$z <- inlabru:::bru_fill_missing(gal1, spp, v$z)
  }
  return(v$z)
}
```

## Modal Fitting for Posterior of $R$ and $n$

## `inlabru`

```{r message=F, warning=F}
spv <- SpatialPoints(v_acs[,3:4]/4300*76)

# construct mesh for INLA-SPDE
v_mesh <- inla.mesh.2d(loc = spv, boundary = region,
                       max.edge = c(80, 800)/4300*76, cutoff = 40/4300*76, max.n.strict = 1300L)

v_spde <- inla.spde2.pcmatern(mesh = v_mesh, alpha = 2,
                              prior.range = c(400/4300*76, 0.5),
                              prior.sigma = c(1.5, 0.5))

# prior
alpha_fun_1 <- function(u){
  qlnorm(pnorm(u), meanlog = 0, sdlog = 0.3)
}

R_fun_1 <- function(u){
  qlnorm(pnorm(u), meanlog = 6.36 - log(4300) + log(76), sdlog = 0.25)
}

# model components
cmp <- coordinates ~ gal1(f.gal1(x,y), model = "offset") +
  beta1(1, model = 'linear') +
  R_internal_1(1, model = 'linear', mean.linear = 0, prec.linear = 1) +
  alpha_internal_1(1, model="linear", mean.linear=0, prec.linear=1) +
  field(coordinates, model = v_spde) + Intercept(1)

# model formula
form <- coordinates ~ Intercept + beta1*exp(-(gal1/R_fun_1(R_internal_1)^2)^alpha_fun_1(alpha_internal_1)) + field

spv <- SpatialPointsDataFrame(spv, data = data.frame(id = 1:nrow(as.data.frame(spv))))

# fit model via inlabru

fit <- lgcp(components = cmp, data = spv,
              samplers = region, domain = list(coordinates = v_mesh), 
              formula = form, options = list(control.inla=list(int.strategy="grid")))

save(fit, file = paste0(output_path, "/UDG_inlabru.rda"))
```

## BOSS

```{r eval = F}
# specify objective function: unnormalized pi(R,n|y)
eval_func <- function(par){
  R1 <- par[1]
  a1 <- par[2]
  
  cmp <- coordinates ~ gal1(f.gal1(x,y), model = "offset")  + 
    beta1(1, model = 'linear') + 
    field(coordinates, model = v_spde) + Intercept(1)
  
  form <- coordinates ~ Intercept + beta1*exp(-(gal1/R1^2)^(a1)) + field
  
  
  fit_inner <- lgcp(components = cmp, data = spv,
                    samplers = region, domain = list(coordinates = v_mesh), 
                    formula = form, options = list(control.inla=list(int.strategy="grid")))
  
  unname(fit_inner$mlik[1,]) + dlnorm(R1, 6.36 - log(4300) + log(76), 0.25, log = T) +
    dlnorm(a1, 0, 0.3, log = T)
}
```


```{r eval = F}
lower = c(2, 0.15)
upper = c(12, 4)

res_opt <- BOSS(eval_func, criterion = 'modal', update_step = 5, max_iter = 100, D = 2,
                        lower = lower, upper = upper,
                        noise_var = 1e-6,
                        modal_iter_check = 10, modal_check_warmup = 20, modal_k.nn = 5,
                        modal_eps = 0.01,
                        initial_design = 5, delta = 0.01^2,
                        optim.n = 5, optim.max.iter = 100)

save(res_opt, file = paste0(output_path, "/UDG_BOSS.rda"))
```

The above BOSS algorithm converged in $60$ iterations.

## Results Comparison

### Marginal Posterior Distributions

```{r}
load(paste0(output_path, "/UDG_inlabru.rda"))

set.seed(1234)
inla.samples.a <- R_fun_1(inla.rmarginal(49500, fit$marginals.fixed$R_internal_1))
inla.samples.b <- alpha_fun_1(inla.rmarginal(49500, fit$marginals.fixed$alpha_internal_1))

load(paste0(output_path, "/UDG_BOSS.rda"))

# construct covariance function for GP regression
data_to_smooth <- list()
unique_data <- unique(data.frame(x = res_opt$result$x, y = res_opt$result$y))
data_to_smooth$x <- as.matrix(dplyr::select(unique_data, -y))
data_to_smooth$y <- (unique_data$y - mean(unique_data$y))

square_exp_cov <- square_exp_cov_generator_nd(length_scale = res_opt$length_scale, signal_var = res_opt$signal_var)

surrogate <- function(xvalue, data_to_smooth, cov){
  predict_gp(data_to_smooth, x_pred = xvalue, choice_cov = cov, noise_var = 1e-6)$mean
}

# grid method to normalize posterior density for plotting joint posterior of (R, n)
ff <- list()
ff$fn <- function(x) as.numeric(surrogate(matrix(x, nrow = 1), data_to_smooth, square_exp_cov))
x.1 <- (seq(from = 2, to = 12, length.out = 300) - 2)/10
x.2 <- (seq(from = 0.15, to = 4, length.out = 300) - 0.15)/3.85
x_vals <- expand.grid(x.1, x.2)
names(x_vals) <- c('x.1','x.2')
x_original <- t(t(x_vals)*(c(12, 4) - c(2, 0.15)) + c(2, 0.15)) 

fn_vals <- apply(x_vals, 1, function(x) ff$fn(matrix(x, ncol = 2)))

# normalize
lognormal_const <- log(sum(exp(fn_vals))*0.02*0.0077*25/9)

post_x <- data.frame(x_original, pos = exp(fn_vals - lognormal_const))


# get posterior samples of (R, n) from BOSS
dx <- unique(post_x$x.1)[2] - unique(post_x$x.1)[1]
dy <- unique(post_x$x.2)[2] - unique(post_x$x.2)[1]
set.seed(123456)
sample_idx <- rmultinom(1:250000, size = 49500, prob = post_x$pos)
sample_x <- data.frame(post_x, n = sample_idx)

samples <- data.frame(do.call(rbind, apply(sample_x, 1, function(x) cbind(runif(x[4], x[1], x[1]+dx), runif(x[4], x[2], x[2] + dy)))))
```

#### Marginal Posterior of $R$

```{r}
# marginal of R
R_marginal <- data.frame(R = c(inla.samples.a, samples[,1]), 
                         method = rep(c('inlabru', 'BOSS'),  c(length(inla.samples.a), length(samples[,1]))))


# plot marginal of R
#tikz(file = "R_marginal_UDG.tex", standAlone=T, width = 4, height = 2)
ggplot(R_marginal, aes(R)) + geom_density(aes(color = method), show_guide = F) +
  stat_density(aes(x = R, colour = method),
                  geom="line",position="identity") + theme_minimal() + xlab('$R$')
#dev.off()
#system('pdflatex R_marginal_UDG.tex')
```

#### Marginal Posterior of $n$

```{r}
# marginal of n
n_marginal <- data.frame(n = c(inla.samples.b, samples[,2]), 
                         method = rep(c('inlabru', 'BOSS'),  c(length(inla.samples.b), length(samples[,2]))))

# plot marginal of n
#tikz(file = "n_marginal_UDG.tex", standAlone=T, width = 4, height = 2)
ggplot(n_marginal, aes(n)) + geom_density(aes(color = method), show_guide = F) +
  stat_density(aes(x = n, colour = method),
                  geom="line",position="identity") + theme_minimal() + xlab('$n$')
#system('pdflatex n_marginal_UDG.tex')
```

From the above, it seems like the marginal distribution of $R$ and $\beta$ are pretty similar from `inlabru` and BOSS. Although, marginal posterior of $R$ seems to be slightly more heavier tailed than that of `inlabru`.

### Joint Posterior Distribution

#### `inlabru`

```{r warning = F}
# get joint posterior sample of (R, n) from inlabru
joint_samp <- inla.posterior.sample(10000, fit, selection = list(R_internal_1 = 1, alpha_internal_1 = 1), seed = 12345)
joint_samp <- do.call('rbind', lapply(joint_samp, function(x) matrix(x$latent, ncol = 2)))

inla.joint.samps <- data.frame(R = R_fun_1(joint_samp[,1]), n = alpha_fun_1(joint_samp[,2]))

# plot joint posterior density of (R,n) from inlabru
#tikz(file = "joint_post_R_n_inlabru.tex", standAlone=T, width = 5, height = 2)
ggplot(inla.joint.samps, aes(R, n)) + stat_density_2d(
  geom = "raster",
  aes(fill = after_stat(density)), n = 300,
  contour = FALSE) +
  geom_point(data = data.frame(R = R_fun_1(fit$summary.fixed$mode[2]), n = alpha_fun_1(fit$summary.fixed$mode[3])), color = 'red', shape = 1, size =0.5) + scale_fill_viridis_c(name = 'Density') + theme_minimal() + xlab('$R$') + ylab('$n$') + xlim(c(2, 12)) + ylim(c(0, 2))
#dev.off()
#system('pdflatex joint_post_R_n_inlabru.tex')
```

#### BOSS

```{r warning = F}
# plot joint posterior of (R, n)
#tikz(file = "joint_post_R_n.tex", standAlone=T, width = 5, height = 2)
ggplot(post_x, aes(x.1,x.2)) + geom_raster(aes(fill = pos)) + 
  geom_point(data = data.frame(x.1 = post_x$x.1[which.max(post_x$pos)], x.2 = post_x$x.2[which.max(post_x$pos)]), color = 'red', shape = 1, size =0.5) + 
  scale_fill_viridis_c(name = 'Density') + theme_minimal() + xlab('$R$') + ylab('$n$') +
  xlim(c(2,12)) + ylim(c(0, 2))
#dev.off()
#system('pdflatex joint_post_R_n.tex')
```

Based on the previous figures, the pictures seem to be more clear: joint posterior of $(R, n)$ based on `inlabru` is again a simple product of the marginal. On the other hand, the results from BOSS indicate that there is in fact a pretty strong correlation structure between $R$ and $n$.

## Posterior Distriution of $\mathcal{U}(s)$

### Posterior distribution of $\mathcal{U}(s)$ and Exceedance probability based on `inlabru`

```{r message = F}
# posterior distribution of spatial random field U through inlabru
U <- predict(fit, fm_pixels(v_mesh, dim = c(300,300), mask = region, format = 'sp'), ~ exp(field), seed = 1234)

set.seed(1108)
sigma <- inla.hyperpar.sample(500000, fit)[,2]
Q <- rnorm(500000, 0, sigma)

q <- quantile(Q, 0.995)

# get exceedance probability through inlabru
exc_q <- excursions.inla(fit, u = q, type = '>', name = 'field', F.limit = 0.6)
sets_q <- continuous(exc_q, v_mesh, 0.05)
proj_q <- inla.mesh.projector(sets_q$F.geometry, dims = c(300, 300))

dat_q <- list()
dat_q$x <- proj_q$x
dat_q$y <- proj_q$y
dat_q$z <- inla.mesh.project(proj_q, field = sets_q$`F`)

r_q <- raster(dat_q)
r_q <- as(r_q, 'SpatialPixelsDataFrame')
r_q <- crop(r_q, region)
r_q <- as.data.frame(r_q)
r_q <- r_q %>%
  mutate(prob = layer) %>%
  dplyr::select(-layer)
```

#### Posterior distribution of $\mathcal{U}(s)$ and Exceedance probability based on BOSS and AGHQ

```{r}
# get posterior of U through BOSS via AGHQ
ff <- list(fn = function(x) {as.numeric(surrogate(matrix(pnorm(x), nrow = 1), data_to_smooth, square_exp_cov)) + sum(dnorm(x, log = TRUE))})
ff$gr = function(x) numDeriv::grad(func = ff$fn, x)
ff$he = function(x) numDeriv::hessian(func = ff$fn, x)

aghq_result = aghq::aghq(ff = ff, 
                         startingvalue = as.numeric(unique_data[which.max(unique_data$y), c(1,2)]), 
                         k = 4)
#optresults = )

###
quad = aghq_result$normalized_posterior$nodesandweights
prob = exp(aghq_result$normalized_posterior$nodesandweights$logpost_normalized) * aghq_result$normalized_posterior$nodesandweights$weights
sampled_freq <- rmultinom(n = 1, size = 49500, prob = prob)
sampled_theta1 = pnorm(quad$theta1) * 10 + 2
sampled_theta2 = pnorm(quad$theta2) * 3.85 + 0.15
sample_theta <- data.frame(R = sampled_theta1, n = sampled_theta2, count = sampled_freq, prob = prob)
```

```{r eval = F}
# compute posterior of U given (R,n)
get_post_U <- function(st){
  R1 <- st[1,'R']
  a1 <- st[1,'n']
  
  cmp <- coordinates ~ gal1(f.gal1(x,y), model = "offset")  +
    beta1(1, model = 'linear') + 
    field(coordinates, model = v_spde) + Intercept(1)
  
  form <- coordinates ~ Intercept + beta1*exp(-(gal1/R1^2)^(a1)) + field
  
  fit_inner <- lgcp(components = cmp, data = spv,
                    samplers = region, domain = list(coordinates = v_mesh), 
                    formula = form, options = list(control.inla=list(int.strategy="grid")))
  
  U <- predict(fit, fm_pixels(v_mesh, dim = c(300,300), mask = region, format = 'sp'), ~ exp(field), seed = 1234)
  U1 <- as.data.frame(U)
  
  set.seed(1108)
  sigma <- inla.hyperpar.sample(500000, fit_inner)[,2]
  Q <- rnorm(500000, 0, sigma)
  
  q <- quantile(Q, 0.995)
  
  exc_0.5 <- excursions.inla(fit_inner, u = q, type = '>', name = 'field', F.limit = 0.6)
  sets_0.5 <- continuous(exc_0.5, v_mesh, 0.05)
  proj_0.5 <- inla.mesh.projector(sets_0.5$F.geometry, dims = c(300, 300))
  
  dat_0.5 <- list()
  dat_0.5$x <- proj_0.5$x
  dat_0.5$y <- proj_0.5$y
  dat_0.5$z <- inla.mesh.project(proj_0.5, field = sets_0.5$`F`)
  
  r_0.5 <- raster(dat_0.5)
  r_0.5 <- as(r_0.5, 'SpatialPixelsDataFrame')
  r_0.5 <- crop(r_0.5, region)
  r_0.5 <- as.data.frame(r_0.5)
  r_0.5 <- r_0.5 %>%
    mutate(prob = layer) %>%
    dplyr::select(-layer)
  
  int <- inla.rmarginal(st[1,'count'], fit_inner$marginals.fixed$Intercept)
  b1 <- inla.rmarginal(st[1,'count'], fit_inner$marginals.fixed$beta1)
  sigma <- inla.hyperpar.sample(st[1,'count'], fit_inner)[,2]
  rho <- inla.hyperpar.sample(st[1,'count'], fit_inner)[,1]
  
  return(list(exc = r_0.5, int = int, b1 = b1, rho = rho, sigma = sigma, U = U1[c('coords.x1', 'coords.x2', 'median')]))
}

BO_aghq_int <- list()

# integrate out (R,n) to get posterior of U through BOSS and AGHQ
for(i in 1:16){
  BO_aghq_int[[i]] <- get_post_U(st = sample_theta[i,])
}

save(BO_aghq_int, file = paste0(output_path, "/UDG_BOSS_AGHQ.rda"))
```

#### Exceedance Probability

```{r}
load(paste0(output_path, "/UDG_BOSS_AGHQ.rda"))
# get exceedance probability through BOSS
mean_exceed_0.5 <- data.frame(BO_aghq_int[[1]][[1]][,c('x','y')])

prob_0.5 <- lapply(lapply(BO_aghq_int, function(x) x$exc), function(y) y$prob)
prob_0.5 <- do.call('rbind', prob_0.5)
prob_0.5 <- colSums(prob_0.5*sample_theta$prob)

mean_exceed_0.5$prob <- prob_0.5

exceed_map <- bind_rows(r_q, mean_exceed_0.5)
exceed_map$method <- rep(c('inlabru', 'BOSS (ours)'), each = 44100)
```

```{r}
# plot the exceedance probability
#options(tikzLatexPackages 
#        =c(getOption( "tikzLatexPackages" ),"\\usepackage{amsfonts}"))
#tikz(file = "exceed_Us_compare.tex", standAlone=T, width = 6, height = 3)
ggplot(exceed_map, aes(x,y)) + geom_raster(aes(fill = prob)) + 
  scale_fill_distiller(palette = 'Spectral', name = 'P(U(s) > C | X)') +
  coord_fixed() + facet_wrap(.~method) +
  xlab('X (kpc)') + ylab('Y (kpc)') + theme_minimal()
#dev.off()
#system('pdflatex exceed_Us_compare.tex')
```

#### Posterior Distribution of $U(s)$

```{r}
# comparison of median of the posterior of U (inlabru vs BOSS)
U_0.5 <- data.frame(BO_aghq_int[[1]]$U[,c('coords.x1','coords.x2', 'median')])

U_med <- lapply(lapply(BO_aghq_int, function(x) x$U), function(y) y$median)
U_med <- do.call('rbind', U_med)
U_med <- colSums(U_med*sample_theta$prob)


U1 <- as.data.frame(U)
U_dat <- bind_rows(U_0.5, U1[c('coords.x1', 'coords.x2', 'median')])
U_dat$method <- rep(c('BOSS (ours)', 'inlabru'), each = 82132)
theBreaks <- c(0, 0.5, 1, 1.5, 2, 3, 4, 5, 6, 7)
theCol = rev(RColorBrewer::brewer.pal(length(theBreaks)-1, 'Spectral'))

# plot it
#tikz(file = "post_med_exp_Us_compare.tex", standAlone=T, width = 5, height = 3)
ggplot(U_dat, aes(coords.x1, coords.x2)) +
  geom_contour_filled(aes(z = median), breaks = theBreaks) +
  scale_fill_manual(values = theCol, name = 'exp(U(s))', guide = guide_legend(reverse = T)) +
  coord_fixed() + facet_wrap(.~method) +
  xlab('X (kpc)') + ylab('Y (kpc)') +
  theme_minimal() +
  theme(legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        strip.background = element_rect(color = NULL, fill = 'white', linetype = 'blank'))
#dev.off()
#system('pdflatex post_med_exp_Us_compare.tex')

```

The final results in terms of the UDG detection is quite similar between `inlabru` and BOSS.




